Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-04-12 msc-publication.tex
add abstract

\abstract{Enter the text of your abstract here.}

@@

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests.  In the horizontal advection test, results of the cubic upwind-biased scheme are compared with the fourth-order centred scheme from \citet{schaer2002}.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ l l l l l l l l l l }
\hline\hline
& \multicolumn{5}{c}{Horizontal advection} & \multicolumn{3}{c}{Terrain following advection} \\
& \multicolumn{2}{c}{Sch\"ar 4th order} & \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{3}{c}{Cubic upwind-biased} \\
& min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 1 & 0 & 0 & 1 \\
BTF 	  & \num{-0.058} & \num{1.001} & \num{0.00767} & \num{-0.0433} & \num{0.928} & \num{0.00154} & \num{-0.0105} & \num{0.983} \\
SLEVE 	  & \num{-0.002} & \num{0.984} & \num{0.00108} & \num{-0.0106} & \num{0.981} & \num{0.0122} & \num{-0.0273} & \num{0.864} \\
Cut cell  & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \num{0.000579} & \num{-0.00670} & \num{0.982} & \num{0.0136} & \num{-0.0297} & \num{0.850} \\
No orography & \num{-0.002} & \num{0.984} & \num{0.000577} & \num{-0.00670} & \num{0.982} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
\hline
\end{tabular}
\end{table*}

@@

The model by \citet{weller-shahrokhi2014} has a Lorenz staggering of variables such that $\theta$ is stored at cell centres and velocity at cell faces.  The inaccuracies in potential temperature become invisible when calculating momentum because $\theta$ is interpolated onto cell faces, so discrete hydrostatic balance is maintained.  This is a manifestation of the Lorenz computational mode, which is a disadvantage of the Lorenz grid that has been widely discussed in the literature \citep{arakawa-konor1996,holdaway2013b}.

\TODO{what might be the physical implications?  1. reduced stability, although not enough to create vertical motion in this instance 2. would disrupt clouds 3. model has no viscosity so thermal mixing should not occur in theory}

@@

\begin{enumerate}
	\item Cut cell grids can be worse than TF grids:
	\begin{itemize}
		\item Lorenz computational mode found on SnapCol grid only (figure~\ref{fig:gw-theta}, \ref{fig:gw-exner-theta})
	\end{itemize}

	\item Cut cell grids can also be better than TF grids in more artificial test cases:
	\begin{itemize}
		\item SnapCol $w$ two orders of magnitude smaller in resting atmosphere test
		\item SnapCol advection test as good as noOrography
	\end{itemize}
\end{enumerate}

\textbf{With the exception of the Lorenz computational mode on the SnapCol grid in the gravity waves test, results were ``satisfactory'' across BTF, SLEVE and SnapCol grids in all three test cases.}

\TODO{Further work: Lorenz computational mode motivates formulation of C-P staggering for cut cell grids}

>>>>

%but are known to suffer from horizontal pressure gradient errors near steep terrain.
\abstract{Terrain following coordinates are widely used in operational models.
More recently, the cut cell method has been proposed as an alternative that can more accurately represent orography.  In order to compare the accuracy of the different grids, it is typically necessary to use different models.  Instead, we use a single model to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  In addition, we formulate a new tracer advection test having a velocity field that is tangential to the basic terrain following coordinate surfaces.
The results of the tracer advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, we find that the Lorenz computational mode is manifested as a vertical zig--zag in potential temperature only on the cut cell grid.
Although errors do vary between grids, overall, a high degree of accuracy is obtained for all tests on all grids.}

@@

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests.  In the horizontal advection test, results of the cubic upwind-biased scheme are compared with the fourth-order centred scheme from \citet{schaer2002}.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ r @{\hspace{2em}} l l @{\hspace{1.5em}} l l l @{\hspace{2em}} l l l }
\hline\hline
& \multicolumn{5}{c}{Horizontal advection} & \multicolumn{3}{c}{Terrain following advection} \\
& \multicolumn{2}{c}{Sch\"ar 4th order} & \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{3}{c}{Cubic upwind-biased} \\
& min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 1 & 0 & 0 & 1 \\
BTF 	  & \num{-0.058} & \num{1.001} & \num{0.00767} & \num{-0.0433} & \num{0.928} & \num{0.00154} & \num{-0.0105} & \num{0.983} \\
SLEVE 	  & \num{-0.002} & \num{0.984} & \num{0.00108} & \num{-0.0106} & \num{0.981} & \num{0.0122} & \num{-0.0273} & \num{0.864} \\
Cut cell  & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \num{0.000579} & \num{-0.00670} & \num{0.982} & \num{0.0136} & \num{-0.0297} & \num{0.850} \\
No orography & \num{-0.002} & \num{0.984} & \num{0.000577} & \num{-0.00670} & \num{0.982} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
\hline
\end{tabular}
\end{table*}

@@

The model by \citet{weller-shahrokhi2014} has a Lorenz staggering of variables such that $\theta$ is stored at cell centres and velocity at cell faces.  The inaccuracies in potential temperature become invisible when calculating momentum because $\theta$ is interpolated onto cell faces, so discrete hydrostatic balance is maintained.  This is a manifestation of the Lorenz computational mode, which is a disadvantage of the Lorenz grid that has been widely discussed in the literature \citep{arakawa-konor1996,holdaway2013b}.  In models that include moist processes, the Lorenz computational mode can disrupt clouds and generate spurious precipitation \citep{hollingsworth1995}.

@@

Cut cell grids reduced errors in two of the four tests.  First, in the horizontal advection test, tracer accuracy on the cut cell grid was almost as good as accuracy on a regular grid with no mountain.  Second, in the resting atmosphere test, spurious vertical velocities were two orders of magnitude smaller on the cut cell grid compared with the terrain folllowing grids.

Conversely, in the terrain following advection test, errors were large on the SLEVE and cut cell grids where velocities were misaligned with the grids.  In the gravity waves test, the Lorenz computational mode was manifested as a zig--zag in potential temperature in the lowest layers in the lee of the mountain.  This spurious error was excited only on the cut cell grid.  This test motivates further work to formulate a Charney--Phillips staggering of variables on cut cell grids.
